Perceptions of Inequality and Meritocracy: Their Interplay in Shaping Preferences for Market Justice in Chile (2016-2023)

### 2025-07-24 resubmission/02-antecedents.qmd
typos

Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023. At the same time, around 55% and 47% of the public between 2016 and 2023 disagreed that effort and talent are rewarded accordingly. This coincides with 90% of citizens agreeing that income differences in Chile are too large [@castillo_changes_2025].
In sum, Chile represents an illustrative case of a particular form of moral economy [@koos_moral_2019], in which socioeconomic inequalities, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.

>>>>

Concerning public opinion, the ELSOC survey shows that Chile has a low level of agreement with the idea that it is fair for those with higher incomes to buy better welfare in the areas of education, healthcare, and pensions, with average support of 8.5% between 2016 and 2023. At the same time, around 55% and 47% of the public between 2016 and 2023 *disagreed* that effort and talent are rewarded accordingly. This coincides with 90% of citizens *agreeing* that income differences in Chile are *too large* [@castillo_changes_2025].

In sum, Chile represents an illustrative case of a particular form of *moral economy* [@koos_moral_2019], in which national socioeconomic characteristics, institutional arrangements, and public attitudes coexist. This suggests that the country combines persistently high levels of economic and social inequality with a strong institutionalization of market-based mechanisms within the welfare system. At the same time, perceptions of inequality are high, and there is widespread public criticism of the commodification of social services. For the above reasons, we argue that the Chilean case represents a relevant context to study the role of perceived income inequality and meritocracy in shaping public preferences regarding the commodification of welfare services, particularly from the perspective of individual change.
